---
title: "Pre-processing"
author: "Mariana Saldarriaga and Ofer Dotan"
date: "2021"
output: html_document
---

# #Load libraries
```{r packages, fig.show='hide', message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(haven)
library(caret)
library(tidyverse) # remember includes dplyr
library(labelled)
library(gtools)
library(car)
library(dplyr)
```

# Load data and remove NAs
```{r}
ud <- read_dta("Uganda ELA Panel wide_Creation.dta", encoding = "latin1")
ud <- remove_labels(ud)
ud <- remove_attributes(ud, "format.stata")
ud[is.na(ud)] = 88
ud[ud == 88] <- NA
ud <- as.data.frame(ud)
```

# Standardization variables 
```{r}
standardize <- function(data) {
  mean_data <- mean(data, na.rm = TRUE)
  sd_data <- sd(data, na.rm = TRUE)
  stand_data <- (data - mean_data)/sd_data
  return(stand_data)
}

ud$z_hsworked_year_empl <- standardize(ud$hsworked_year_empl)
ud$z_hsworked_year_self <- standardize(ud$hsworked_year_self)
ud$z_Qhsworked_year_empl <- standardize(ud$Qhsworked_year_empl)
ud$z_Rhsworked_year_empl <- standardize(ud$Rhsworked_year_empl)
ud$z_satisfaction_income <- standardize(ud$satisfaction_income) 
ud$z_study_hours <- standardize(ud$study_hours) 
ud$z_Qstudy_hours <- standardize(ud$Qstudy_hours) 
ud$z_Rstudy_hours <- standardize(ud$Rstudy_hours) 
ud$z_age_Imarry <- standardize(ud$age_Imarry) 
ud$z_Rage_Imarry <- standardize(ud$Rage_Imarry) 
ud$z_M_marrywho <- standardize(ud$M_marrywho)
```

# First subset - including endline and follow-up variables + indexes
```{r pressure, echo=FALSE}
ud_subset <- ud %>% 
  subset(select = c("Expenditure_totDF", "QExpenditure_totDF", "RExpenditure_totDF", "HHAssetvalue_total", "z_Qhsworked_year_empl", "HHF_loanbrac", "HHM_whoshouldearn","M_ablework_if", "below16", "z_satisfaction_income", 
                    "M_children", "M_marrywhen", "M_marrywho",
                    "M_wanttowork_if","QC_clubheard","Qincome_year_ind", "Rincome_year_ind", 
                    "QC_clubparticipateIMP", "QC_stillgoing", "QE_Denrolled", "Qage", 
                    "Qback_school","Qhsworked_year_empl","Qhsworked_year_self", "Qincind_empl",
                    "Qincind_selfempl", "Qincome_year_ind", "QlifeskillMOREfew",
                    "QlifeskillMOREfewIMP","QlivelihoodMOREfew", "QlivelihoodMOREfewIMP", "Qoften1WEEK",
                    "Qoften3WEEK", "z_hsworked_year_empl", "Qstudy_hours", "Qworry_job",
                    "RC_clubheard","RC_clubparticipateIMP","RC_stillgoing","RE_Denrolled", 
                    "Rage", "Rback_school", "Rhsworked_year_self", "Rhsworked_year_empl", "Rincind_empl",
                    "Rincind_selfempl", "Rstudy_hours",
                    "Rworry_job", "age", "back_school","baseline", "branch_name", "branchno",
                    "dist_nearclub","endline", "follow_up", "hsworked_year_empl", "hsworked_year_self", "id",
                    "idno","incind_empl","incind_selfempl","income_year_ind", "lifeskillMOREfew",
                    "lifeskillMOREfewIMP","livelihoodMOREfew","livelihoodMOREfewIMP",
                    "often1WEEK","often3WEEK", "satisfaction_income", "selfempl","study_hours",
                    "treatment", "villid", "worry_job", "age_Imarry", "Rage_Imarry",          
                    "rural", "rich","z_QEntrep_total", "z_Qany_iga", "z_Qselfempl",
                    "z_Qempl","z_QExpenditure_totDF","QM_chi","Qpart", "z_QM_chi","z_Qpart","z_QR_sexu","z_Qsex_p",
                    "z_QRhiv_s","z_Qalways_c", "z_Qother_c", "QM_baby_no",           
                    "QM_son", "z_Qempowerment", "z_QM_i_ageF","z_QM_i_ageM","z_QM_baby_no","z_QM_baby_ageF",       
                    "z_QM_daught","z_QM_son","Qcontrol_body","Qaspiration","Qiga","z_REntrep_total","z_Rany_iga",
                   "z_Rempl", "z_RExpenditure_totDF", "RM_chi", "Rpart", "RR_sexu", "Ralways_c","Rother_c",
                    "z_RM_chi", "z_Rpart", "z_RR_sexu", "z_Rsex_p", "z_RRhiv_s", "z_Ralways_c","z_Rother_c",
                    "z_Rempowerment","z_RM_i_ageF","z_RM_i_ageM","z_RM_baby_no","z_RM_baby_ageF",
                    "z_RM_daught","z_RM_son", "Rcontrol_body","Raspiration","Riga", "z_Entrep_total","z_any_iga",
                    "z_selfempl","z_empl","z_Expenditure_totDF",
                    "zALL_Entrep_total","zALL_any_iga","zALL_selfempl","zALL_empl",
                    "zALL_Expenditure_totDF", "M_chi","part",                
                    "always_c","other_c","z_M_chi",
                    "z_part","z_R_sexu","z_sex_p","z_Rhiv_s","z_always_c","z_other_c","zALL_R_sexu",
                    "zALL_sex_p","zALL_Rhiv_s","zALL_always_c","zALL_other_c",
                    "z_empowerment","z_M_i_ageF","z_M_i_ageM","z_M_baby_no",
                    "z_M_baby_ageF","z_M_daught","z_M_son","zALL_empowerment",
                    "zALL_M_i_ageF","zALL_M_i_ageM", "z_hsworked_year_self",
                    "zALL_M_baby_no","zALL_M_baby_ageF","zALL_M_daught","zALL_M_son","control_body","aspiration", 
                    "iga","igaALL", "_Bbranch_na_2","_Bbranch_na_3","_Bbranch_na_4", "_Bbranch_na_5",
                    "_Bbranch_na_6","_Bbranch_na_7","_Bbranch_na_8","_Bbranch_na_9", "_Bbranch_na_10", "Attitude",
                    "E_Denrolled", "empl", "partner", "empowerment","Qempowerment","Rempowerment")
         ) 
```


# Rename variables and keep only panel (baseline and follow-up)
```{r}
ud_subset <- ud_subset %>% 
  rename(expenditure_goods = Expenditure_totDF,
         foll_expend_goods = RExpenditure_totDF,
         endline_expend_goods = QExpenditure_totDF,
         endline_controlbody = Qcontrol_body,
         endline_aspiration = Qaspiration,
         endline_income_gen_act = Qiga,
         endline_empowerment_stand = z_Qempowerment,
         treatment = treatment, ## baseline
         rich = rich, 
         value_assets = HHAssetvalue_total, # not make assumption, model predict
         below16 = below16, # transformation age
         age = age, # keep original values; control in paper
         branchno = branchno, # control in paper i.branch_name
         rural = rural,
         vill_id = villid, # cluster paper
         want_respect = Attitude,
         enroll_school = E_Denrolled,
         back_school = back_school,
         dist_nearclub = dist_nearclub,
         empl = empl,
         loan_brac = HHF_loanbrac, 
         att_earn_moneyfam = HHM_whoshouldearn,
         ablework_married = M_ablework_if,
         who_decidehusband = M_marrywho, 
         work_married = M_wanttowork_if, 
         index_income_gen_act = iga, # z_Entrep_total,z_any_iga,z_selfempl,z_empl,z_Expenditure_totDF
         index_aspiration = aspiration, # z_M_i_ageF,z_M_i_ageM,z_M_baby_no,z_M_baby_ageF,z_M_daught,z_M_son
         index_controlbody = control_body, # z_M_chi,z_part,z_R_sexu,z_sex_p,z_Rhiv_s,z_always_c,z_other_c
         children_stand = z_M_chi, # -M_children(reverse), standardize dummy --> M_chi # part control over body index
         children = M_children,
         partner = partner,
         partner_stand = z_part, # -partner, standardize dummy -->part # part control over body index
         sex_unwilling = z_R_sexu, # -R_sexunwilling, standardize dummy --> R_sexu # part control over body index
         index_empowerment_stand = z_empowerment, # Not in aspiration index!!!!
         index_empowerment = empowerment,
         back_school = back_school, 
         income = income_year_ind, # already trimmed
         life_skills = lifeskillMOREfewIMP, 
         livelihood = livelihoodMOREfewIMP,
         club_sometimes = often1WEEK, 
         club_frequently = often3WEEK,
         satisf_income = satisfaction_income,
         study_hours = study_hours, 
         worry_job = worry_job, # we should inverse it 1 is not worry and this is bad, normally 1 is good
         branch_name = branch_name, # paper uses i.branch_name as control variable --> base _Bbranch_na_1 #
         endline_income = Qincome_year_ind, 
         foll_income = Rincome_year_ind,
         foll_club_heard = RC_clubheard, ## follow up
         foll_club_part = RC_clubparticipateIMP, # imputed values
         foll_still_goclub = RC_stillgoing,
         foll_enroll_school = RE_Denrolled, 
         foll_age = Rage,
         foll_back_school = Rback_school,
         foll_study_hours = Rstudy_hours,
         foll_worry_job = Rworry_job,
         foll_children = z_RM_chi, # standardize -RM_Children (dummy), # part control over body index
         foll_partner = z_Rpart, # standardize -Rpartner (dummy) # part control over body index
         foll_sex_unwilling = z_RR_sexu, # standardize -RR_sexunwilling (dummy) # part control over body index
         foll_index_income_gen_act = Riga, # z_REntrep_total,z_Rany_iga,z_Rselfempl,z_Rempl,z_RExpenditure_totDF
         foll_index_controlbody = Rcontrol_body, # z_RM_chi,z_Rpart,z_RR_sexu,z_Rsex_p,z_RRhiv_s,z_Ralways_c z_Rother_c
         foll_index_aspiration = Raspiration, # z_RM_i_ageF,z_RM_i_ageM,z_RM_baby_no,z_RM_baby_ageF,z_RM_daught,z_RM_son
         foll_index_empowerment_stand = z_Rempowerment, # Not in aspiration index!!!!
         foll_index_empowerment = Rempowerment,
         endline_empowerment = Qempowerment,
         foll_expend_goods_stand = z_RExpenditure_totDF
  )
```

# Final subset
```{r}
df.uganda <- as.data.frame(ud_subset) %>%
  select(foll_expend_goods, endline_expend_goods, expenditure_goods, endline_income,foll_income, endline_controlbody,endline_aspiration,endline, index_empowerment_stand, endline_income_gen_act,endline_empowerment,baseline, value_assets, ablework_married, children, who_decidehusband, work_married, age, back_school, branch_name, branchno, dist_nearclub, follow_up, z_hsworked_year_self, z_hsworked_year_empl, id, income, life_skills, livelihood, club_sometimes, club_frequently, satisf_income, z_selfempl, study_hours, treatment, vill_id, worry_job, rural, rich, below16, sex_unwilling, foll_index_controlbody, foll_index_aspiration, foll_index_income_gen_act, z_Entrep_total, z_any_iga, z_empl, z_Expenditure_totDF, partner, empl, index_empowerment, index_controlbody, index_aspiration, index_income_gen_act, want_respect, enroll_school, foll_expend_goods_stand, foll_index_empowerment_stand, foll_index_empowerment, endline_empowerment, endline_empowerment_stand, "_Bbranch_na_2", "_Bbranch_na_3", "_Bbranch_na_4", "_Bbranch_na_5", "_Bbranch_na_6", "_Bbranch_na_7", "_Bbranch_na_8", "_Bbranch_na_9", "_Bbranch_na_10")

```

# Means outcomes (follow-up and endline)
```{r}
df.uganda$post_income_gen_act <- rowMeans(df.uganda[,c("endline_income_gen_act", "foll_index_income_gen_act")], na.rm=TRUE)

df.uganda$post_income_gen_act_max <- pmax(df.uganda$endline_income_gen_act, df.uganda$foll_index_income_gen_act, na.rm = TRUE)
                                    
df.uganda$post_control_body <- rowMeans(df.uganda[,c("endline_controlbody", "foll_index_controlbody")], na.rm = TRUE)

df.uganda$post_control_body_max <- pmax(df.uganda$endline_controlbody, df.uganda$foll_index_controlbody, na.rm = TRUE)

df.uganda$post_empowerment <- rowMeans(df.uganda[,c("endline_empowerment", "foll_index_empowerment")], na.rm = TRUE)
  
df.uganda$post_empowerment_max <- pmax(df.uganda$endline_empowerment, df.uganda$foll_index_empowerment, na.rm = TRUE)

df.uganda$post_aspirations <- rowMeans(df.uganda[,c("endline_aspiration", "foll_index_aspiration")], na.rm = TRUE)

df.uganda$post_aspirations_max <- pmax(df.uganda$endline_aspiration, df.uganda$foll_index_aspiration, na.rm = TRUE)

df.uganda$post_income <- rowMeans(df.uganda[,c("endline_income", "foll_income")], na.rm = TRUE)

df.uganda$post_income_max <- pmax(df.uganda$endline_income, df.uganda$foll_income, na.rm = TRUE)

df.uganda$post_expend <- rowMeans(df.uganda[,c("endline_expend_goods", "foll_expend_goods")], na.rm = TRUE)
```

# Create a factor variable using the quantiles of continuous variables
```{r}
df.uganda$age_quant <- quantcut(df.uganda$age, q=3, na.rm=TRUE)
#df.uganda$income_quant <- quantcut(df.uganda$income, q=3, na.rm=TRUE, names = TRUE) #not working
df.uganda$study_hours_quant <- quantcut(df.uganda$study_hours, q=3, na.rm=TRUE)
df.uganda$index_aspiration_quant <- quantcut(df.uganda$index_aspiration, q=4, na.rm=TRUE)
df.uganda$index_iga_quant <- quantcut(df.uganda$index_income_gen_act, q=4, na.rm=TRUE)
df.uganda$index_empowerment_quant <- quantcut(df.uganda$index_empowerment, q=3, na.rm=TRUE)
df.uganda$index_controlbody_quant <- quantcut(df.uganda$index_controlbody, q=3, na.rm=TRUE)
df.uganda$expend_quant <- quantcut(df.uganda$expenditure_goods, q=3, na.rm=TRUE)
```

# Recode categorical variables as dummies
```{r}
# Dummy now - we need to specify this was recoded; any level of satisfaction coded as satisfied)
df.uganda$satisf_income <- car::recode(df.uganda$satisf_income, "c('0') = '0'; c('1','2','3','4','5','6')='1'")

# Recoded taking 1:5 (354 obs) doesn't want respect and 6:10 (5516) does want respect 
df.uganda$want_respect <- car::recode(df.uganda$want_respect, "c('1','2','3','4','5') = '0'; c('6','7','8','9','10') ='1'")
```

# Clean data set
```{r}
df.uganda <- df.uganda %>% 
 mutate(branchno = as.factor(branchno),
        treatment = as.factor(treatment),
        value_assets = as.numeric(value_assets),
        work_married = as.factor(work_married),
        enroll_school = as.factor(enroll_school),
        age = as.numeric(age),
        who_decidehusband = as.factor(who_decidehusband),
        children = as.factor(children),
        back_school = as.factor(back_school),
        branchno = as.factor(branchno),
        dist_nearclub = as.numeric(dist_nearclub),
        z_hsworked_year_empl = as.numeric(z_hsworked_year_empl),
        z_hsworked_year_self = as.numeric(z_hsworked_year_self),
        income = as.numeric(income),
        study_hours = as.numeric(study_hours),
        rich = as.factor(rich),
        below16 = as.factor(below16),
        foll_expend_goods = as.numeric(foll_expend_goods),
        endline_expend_goods = as.numeric(endline_expend_goods),
        expenditure_goods = as.numeric(expenditure_goods),
        foll_expend_goods_stand = as.numeric(foll_expend_goods_stand),
        foll_index_empowerment = as.numeric(foll_index_empowerment),
        foll_index_empowerment_stand = as.numeric(foll_index_empowerment_stand),
        foll_index_controlbody = as.numeric(foll_index_controlbody), 
        foll_index_aspiration = as.numeric(foll_index_aspiration),
        foll_index_income_gen_act = as.numeric(foll_index_income_gen_act),
        ablework_married = as.factor(ablework_married),
        satisf_income = as.factor(satisf_income),
        life_skills = as.factor(life_skills),
        club_sometimes = as.factor(club_sometimes),
        club_frequently =as.factor(club_frequently),
        livelihood = as.factor(livelihood),
        vill_id = as.factor(vill_id),
        rural = as.factor(rural),
        treatment = as.factor(treatment),
        worry_job = as.factor(worry_job),
        z_Entrep_total = as.numeric(z_Entrep_total),
        sex_unwilling = as.factor(sex_unwilling),
        index_empowerment = as.numeric(index_empowerment),
        z_any_iga = as.factor(z_any_iga),
        z_empl = as.factor(z_empl),
        z_selfempl = as.factor(z_selfempl),
        z_Expenditure_totDF = as.numeric(z_Expenditure_totDF),
        partner = as.factor(partner),
        index_controlbody = as.numeric(index_controlbody),
        index_aspiration = as.numeric(index_aspiration),
        index_income_gen_act = as.numeric(index_income_gen_act),
        want_respect = as.factor(want_respect),
        enroll_school = as.factor(enroll_school),
        branch_name = as.factor(branch_name),
        age_quant = as.factor(age_quant),
        index_aspiration_quant = as.factor(index_aspiration_quant),
        index_iga_quant = as.factor(index_iga_quant),
        index_empowerment_quant = as.factor(index_empowerment_quant),
        index_controlbody_quant = as.factor(index_controlbody_quant),
        empl = as.factor(empl), 
        study_hours_quant = as.factor(study_hours_quant), 
        expend_quant = as.factor(expend_quant)
        )

```

# NAs
```{r}
df.uganda[df.uganda == NaN] <- NA
```

# Save final data
```{r}
save(df.uganda, file = "df.uganda.RData")
```
