---
title: "Causal Random Forest . Outcome: Expenditure on Goods [UGX]"
author: "Ofer Dotan and Mariana Saldarriaga"
date: "2021"
output: html_document
---

# Load libraries
```{r packages, fig.show='hide', message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(grf)
library(haven)
library(aod)
library(caret)
library(tidyverse) 
library(expss)
library(base)
library(readr)
library(stats)
library(knitr)
library(kableExtra)
library(gtools)
```

# Load data
```{r}
load("df.uganda.RData")
```

```{r}
set.seed(1236)
options(na.action = 'na.pass')
```

# Generate X(covariates), Y(outcome), W (treatment), C(cluster) matrixes
```{r, include=FALSE}
df.uganda <- as.data.frame(df.uganda) %>%
  filter(complete.cases(vill_id, foll_expend_goods))

Y.outcome <- as.matrix(df.uganda$foll_expend_goods)


#Baseline variables
X.vars <- stats::model.matrix(~ children + 
                            branchno + 
                            age_quant +
                            expenditure_goods + 
                            partner +
                            enroll_school +
                            rural+
                            rich+
                            club_frequently + 
                            club_sometimes + 
                            empl+
                            study_hours_quant +
                            worry_job +
                            work_married+
                            ablework_married+
                            want_respect+
                            who_decidehusband+
                            index_aspiration_quant+
                            index_iga_quant+
                            index_empowerment_quant+
                            satisf_income,
                            data= df.uganda)

#Cluster
C.vill.id <- as.numeric(df.uganda$vill_id)

#Treatment assignment
W.treatment <- as.numeric(levels(df.uganda$treatment)[as.integer(df.uganda$treatment)])
```

# Estimate expected responses(Y.hat) and treatment propensities(W.hat) with regression forests
```{r}
Y.forest <- regression_forest(X.vars, 
                             Y.outcome, 
                             clusters = C.vill.id, 
                             equalize.cluster.weights = TRUE)
Y.hat <- predict(Y.forest)$predictions

W.forest <- regression_forest(X.vars, 
                             W.treatment, 
                             clusters = C.vill.id,
                             equalize.cluster.weights = TRUE)
W.hat <- predict(W.forest)$predictions
```

```{r}
# Alternative option tu manually turn on the parameters
#params <- tune_causal_forest(X.vars, Y.outcome, W.treatment, Y.hat, W.hat, C.vill.id, equalize.cluster.weights = TRUE, tune.num.draws = 10000, tune.num.trees = 10000)$params
```

# Train raw causal forest (no important variable selection)
```{r}
cf.raw = causal_forest (X.vars, 
                        Y.outcome, 
                        W.treatment,
                        num.trees = 20000,
                        Y.hat = Y.hat, 
                        W.hat = W.hat,
                        clusters = C.vill.id,
                        equalize.cluster.weights = TRUE,
                        tune.parameters = "all",
                        orthog.boosting = TRUE,
                        #min.node.size = as.numeric(params["min.node.size"]),
                        #mtry = as.numeric(params["mtry"]),
                        #alpha = as.numeric(params["alpha"]),
                        #imbalance.penalty = as.numeric(params["imbalance.penalty"]),
                        honesty = TRUE)
```

# Find important variables and plot them
```{r causal_forest_varimp_table, results="asis", message=FALSE, echo=FALSE}
var_imp <- variable_importance(cf.raw)
selected.idx <- which(var_imp > mean(var_imp))

#Table variable importance
names <- colnames(X.vars[,selected.idx])
sort_var_imp <- sort(selected.idx, decreasing=TRUE) 
TableE.1 <-  as.data.frame(sort_var_imp, row.names = names) %>%  # names no value var_imp!!!
knitr::kable("html", digits = 4, row.names = TRUE, caption = "Table E.1: List of Selected Variables. Outcome: Expenditure on Goods[UGX]")  %>% kable_classic("hover", full_width = T, html_font = "\"Times New Roman\"") %>% kable_styling(bootstrap_options=c("striped", "hover", "condensed","responsive"), full_width=FALSE)%>% footnote("Causal forest determines variable importance by a weighted sum capturing the number of splits a feature had in every depth layer in the model. Presented in descending order are variables that were above the mean weight of all the variables used for splitting in the model. 'c.") 

save_kable(TableE.1, "TableE.1.png")
TableE.1
```

# Train a second causal forest using only important variables index
```{r}
cf <- causal_forest(X.vars[,selected.idx], 
                      Y.outcome, W.treatment, 
                      Y.hat = Y.hat, 
                      W.hat = W.hat, 
                      num.trees = 20000,
                      clusters = C.vill.id,  
                      orthog.boosting = TRUE,
                      equalize.cluster.weights = TRUE,
                      tune.parameters = "all",
                      #min.node.size = as.numeric(params["min.node.size"]),
                      #sample.fraction = as.numeric(params["sample.fraction"]),
                      #mtry = as.numeric(params["mtry"]),
                      #alpha = as.numeric(params["alpha"]),
                      #imbalance.penalty = as.numeric(params["imbalance.penalty"]),
                      honesty= TRUE)
```

# Estimate ATE for cf model with important variables
```{r}
# Average treatment effect for the model with important variables
ATE.cf <- as.data.frame(average_treatment_effect(cf))
saveRDS(ATE.cf, file = "ATE.expend.rds") 

# Take objects of ATE from other outcomes
#ate <- as.data.frame(cbind(ATE.expend, ATE.empower, ATE.cob))

# Table ATE for the 3 outcomes
rownames(ate) <- c("Estimated ATE", "Standard error")
knitr::kable(ate, col.names = c("Expenditure on Goods","Gender Empowerment", "Control Over Body")) %>% kable_styling() %>% kable_classic("hover", full_width = T) %>% kable_classic("hover", full_width = T,html_font = "\"Times New Roman\"") %>% footnote("Estimated ATEs by a trained causal forest using only baseline variables selected by theory for the linear models. The outcome Expenditure codes consumption of goods in several goods categories. The variable is measured in UGX currency. Exchange rate was 1,700UGX/1$. The Gender Empowerment index cummulates answers regarding gender roles and it is rescaled from 0 to 100 where higher values indicate more empowerment. The index COB is composed by z-score standardised variables. It captures the extent of reproductive health literacy and family planning where higher values mean more COB. See Table 1 in Methodology chapter for the complete specification of these outcomes.", fixed_small_size = TRUE)
```

# Estimate treatment effects (tau.hat --> Athey CATE) using out-of-bag prediction 
```{r}
tau.hat <- predict(cf, num.trees = 20000, estimate.variance = TRUE)

#Histogram
tau.hat.hist <- ggplot(tau.hat, aes(predictions)) + geom_histogram(color="#52854C", fill="#52854C", alpha=.4, bins = 35)

tau.hat.hist +labs (title = "Histogram of out-of-bag CATE predictions from trained causal forest", subtitle = "Expenditure on goods in the last month[UGX]", caption = "Note. Predictions made on held-out data from the full sample that was not used for training.") + ylab("Frequency") + 
   xlab("Predictions CATE (in UGX currency)") +  theme_bw(base_family = "Times New Roman")+ theme(plot.caption = element_text(hjust = 0))
```

# Estimate CATE and plot estimations
```{r, fig.align='center'}

source("get_plot_data.R")

X.vars.cate <- df.uganda %>% select(
                            Rural = rural,
                            Expenditure = expend_quant,
                            Age = age_quant,
                            Partner = partner,
                            "Empowerment" = index_empowerment_quant,
                            Children = children,
                            #Enrolled = enroll_school,
                            #Rich = rich,
                            "Study hours" = study_hours_quant)

PLOT_DATA <- get_plot_data(cf, X.vars.cate)

# visualize these
plot.cates <- ggplot(PLOT_DATA, aes(x = level, y = estimate, colour = name)) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = .2) +
  geom_hline(yintercept = 0, linetype = 3) +
  theme_light() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.text.y = element_text(colour = "black"),
    strip.background = element_rect(colour = NA, fill = NA),
    legend.position = "none"
    ) + 
  facet_grid(name ~ ., scales = "free_y") +
  coord_flip() +
  scale_y_continuous(name="Estimate CATE by Subgroups")

plot.cates+ggtitle("Heterogeneity Across Covariates", subtitle = "Outcome:Expenditure on goods in the last month[UGX]") + labs(caption = "Notes: 95% confidence intervals are calculated based on the standard error of each covariate level estimated ATE effect which can be seen by the colored dots.
The black dotted line show the boundary for a zero impact.
For expenditure variable, the goods categories are jewelry/ornaments, cosmetics/makeup, clothes, hairdressers, shoes/footwear, going to restaurants/bars/teashop/cafe, 
talk time for your mobile phone and presents/gifts.The top 1% outliers of the expenditure variable have been removed. copied from Bandiera et al., 2014") + theme(plot.caption = element_text(hjust = 0))
```

# Heterogeneity across subgroups (Athey, 2019)
```{r}
# Athey, S. (2019). Estimation of Heterogeneous Treatment Effects and Optimal Treatment Policies [Rmarkdown].
```

```{r  causal_forest_subgroup_het1}
# Manually creating subgroups n-tiles
num_tiles <- 3  
df.uganda$cate <- tau.hat$predictions
df.uganda$ntile <- factor(ntile(tau.hat$predictions, n=num_tiles))
```

```{r sample_ate}
# Average treatment effects within subgroups
ols_sample_ate <- lm("Y.outcome ~ ntile + ntile:W.treatment", data=df.uganda)
estimated_sample_ate <- coef(summary(ols_sample_ate))[(num_tiles+1):(2*num_tiles), c("Estimate", "Std. Error")]
hypothesis_sample_ate <- paste0("ntile1:W.treatment = ", paste0("ntile", seq(2, num_tiles), ":W.treatment"))
ftest_pvalue_sample_ate <- linearHypothesis(ols_sample_ate, hypothesis_sample_ate)[2,"Pr(>F)"]
```

```{r  ate_combined_table, results="asis", message=FALSE, echo=FALSE}
# Round the estimates and standard errors before displaying them
estimated_sample_ate_rounded <- round(signif(estimated_sample_ate, digits = 6), 6)

# Format Table: Parenthesis, row/column names
sample_ate_w_se <- c(rbind(estimated_sample_ate_rounded[,"Estimate"], paste0("(", estimated_sample_ate_rounded[,"Std. Error"], ")")))

table <- cbind("Sample ATE" = sample_ate_w_se)

table <- rbind(table, round(signif(c(ftest_pvalue_sample_ate), digits = 5), 4)) # add p-value to table
left_column <- rep('', nrow(table))
left_column[seq(1, nrow(table), 2)] <-
    cell_spec(c(paste0("ntile", seq(num_tiles)), "P-Value"),
              format = "html", escape = FALSE, color = "black", bold = TRUE)
table <- cbind(" " = left_column, table)

# Output table
table %>%
  kable("html", escape = FALSE, row.names = FALSE, caption = "Estimated CATE by Ntiles. Outcome: Expenditure on Goods [UGX]") %>% kable_classic("hover", full_width = F) %>%
  kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE) %>% kable_classic("hover", full_width = T,html_font = "\"Times New Roman\"") %>%
  footnote(general = "Average treatment effects per subgroup (ntile) defined by out-of-bag CATE.<br>
           P-value is testing <i>H<sub>0</sub>: ATE is constant across ntiles</i>.<br>
           Sample ATE uses an F-test. For full description of outcome see Table 1. Methodology chapter.",
           escape=FALSE)
```

```{r ate_combined_plot, echo=FALSE, results='as_is', fig.height=4}
# Transform to data tables with relevant columns
estimated_sample_ate <- as.data.frame(estimated_sample_ate)
estimated_sample_ate$Method <- "Sample ATE. Outcome:Expenditure on Goodsin the Last Month [UGX]"
estimated_sample_ate$Ntile <- as.numeric(sub(".([0-9]+).", "\\1", rownames(estimated_sample_ate)))

colnames(estimated_sample_ate) <- c("Estimate", "SE", "Method", "Ntile")

# plot
ggplot(estimated_sample_ate) +
  geom_pointrange(aes(x = Ntile, y = Estimate, ymax = Estimate + 1.96 * SE, ymin = Estimate - 1.96 * SE, color = Method), 
                  size = 0.5,
                  position = position_dodge(width = .5)) +
  geom_errorbar(aes(x = Ntile, ymax = Estimate + 1.96 * SE, ymin = Estimate - 1.96 * SE, color = Method), 
                width = 0.4,
                size = 0.75,
                position = position_dodge(width = .5)) +
  theme_bw(base_family = "Times New Roman") +
  labs(x = "N-tile", y = "ATE Estimate", title = "ATE within N-tiles (as defined by predicted CATE)",
       subtitle = "Outcome: Expenditure on Goods[UGX]",
       caption = "Note. 95% CI")
```

```{r pairwise_ntile_ate_test, results='hide', message=FALSE, echo=TRUE, warning=FALSE, error=FALSE}
# We test if the average treatment effect is different between all two pairs of n-tiles (p-value not multiple testing)
p_values_tile_by_tile <- matrix(nrow = num_tiles, ncol = num_tiles)
differences_tile_by_tile <- matrix(nrow = num_tiles, ncol = num_tiles)
stderror_tile_by_tile <- matrix(nrow = num_tiles, ncol = num_tiles)
hypotheses_grid <- combn(1:num_tiles, 2)

invisible(apply(hypotheses_grid, 2, function(x) {
  .diff <- with(estimated_sample_ate, Estimate[Ntile == x[2]] - Estimate[Ntile == x[1]])
  .se <- with(estimated_sample_ate, sqrt(SE[Ntile == x[2]]^2 + SE[Ntile == x[1]]^2))

  differences_tile_by_tile[x[2], x[1]] <<- .diff
  stderror_tile_by_tile[x[2], x[1]] <<- .se
  p_values_tile_by_tile[x[2], x[1]] <<- 1 - pnorm(abs(.diff/.se)) + pnorm(-abs(.diff/.se))
}))
```

```{r echo=FALSE, results='asis'}
# Little trick to display p-values under mean difference values in HTML
diffs <- matrix(nrow = num_tiles, ncol = num_tiles)
invisible(apply(hypotheses_grid, 2, function(x) {
  d <- differences_tile_by_tile[x[2], x[1]]
  s <- stderror_tile_by_tile[x[2], x[1]]
  p <- p_values_tile_by_tile[x[2], x[1]]
  top <- cell_spec(round(d, 3), "html",
            background=case_when(is.na(p) || (p > 0.05) ~ "white",
                                 p >= 0.1               ~ "gray",
                                 TRUE                   ~ "black"),
            color=ifelse(is.na(p), "white", ifelse(p < 0.1, "white", "gray")))
  value <- ifelse(is.na(p), "", paste0(top, " <br> ", "(", round(s, 3), ")"))
  diffs[x[2], x[1]] <<- value
}))

diffs <- as.data.frame(diffs) %>% mutate_all(as.character)
rownames(diffs) <- paste0("tile", 1:num_tiles)
colnames(diffs) <- paste0("tile", 1:num_tiles)

# Title of table
caption <- "Table 9: Pairwise n-tile differences:<br>
ATE differences between tile 1 and tile 3 for outcome: Expenditure on Goods [UGX]"

# Styling color and background
color <-function(x) ifelse(is.na(x), "white", "gray")

diffs %>%
  rownames_to_column() %>%
  mutate_all(function(x) cell_spec(x, "html", escape=FALSE, color=color(x))) %>%
  kable(format="html", caption=caption, escape = FALSE) %>% kable_classic("hover", full_width = F) %>% 
  kable_styling(bootstrap_options=c("condensed", "responsive"), full_width=FALSE) %>% kable_classic("hover", full_width = T,html_font = "\"Times New Roman\"") %>%
  footnote(general='Standard errors in parenthesis. Significance:
<ul>
  <li>No background color: p ≥ 0.05
  <li><span style="color: white;border-radius: 4px; padding-right: 4px; padding-left: 4px; background-color: gray;">Gray</span> background: p < 0.05
  <li><span style="color: white;border-radius: 4px; padding-right: 4px; padding-left: 4px; background-color: black;">Black</span> background: p < 0.01
</ul>
', escape=F)
```

# Diagnostics Causal Forest

## Best linear predictor analysis to understand if the model detects heterogeneity
```{r}
tst <- test_calibration(cf)

stargazer(tst, type = "text", dep.var.caption = "Estimate",  title = "Best linear fit. Model Expenditure", notes = "Best linear fit using predictions (on held-out data) as well as the mean forest prediciton as regressors, along with one-sided heteroskedasticity-robust (HC3) SEs", notes.align = "l")
```

## Validating the overlap assumption 
```{r}
propensity.forest <- regression_forest(X.vars[,selected.idx], W.treatment)
W.hat.p <- predict(propensity.forest)

overlap <- ggplot(W.hat.p, aes(predictions)) + geom_histogram(color="#00AFBB", fill="#00AFBB", alpha=.4, bins=35)


overlap+labs(title = "Expenditure",
              subtitle = "Assumption holds if observations are bounded away from 0 and 1")+
   ylab("Predictions") + xlab("Propensity score") + theme_bw()

# Alternative method to assess overlap
#hist(tau.hat <- cf$W.hat)
```

## Estimate standard errors for test data and plot them
```{r}
standard.error = sqrt(tau.hat$variance.estimates) # This is the sigma part
hist(standard.error, col = "#00AFBB", main = "Histogram of estimated standard error of causal forest", sub = "Outcome: Expenditure ",xlab="Standard error")
```

## Estimate model bias
```{r}
p <- mean(W.treatment)
Y.hat.0 <- cf$Y.hat - W.hat * tau.hat
Y.hat.1 <- cf$Y.hat + (1 - W.hat) * tau.hat
bias <- (W.hat - p) * (p * (Y.hat.0 - mean(Y.hat.0))  + (1 - p) * (Y.hat.1 - mean(Y.hat.1)))
hist(bias$variance.estimates / sd(Y.outcome), col = "#00AFBB", main = "Model bias", sub = "Outcome: Expenditure", xlab = "bias/sd(Y)")
```

## Causal Forest Predictions
```{r causal_forest_oob_pred_table, results="asis", message=FALSE, echo=FALSE}
# The first few rows of the cf output look like this...
kable_styling(kable(head(tau.hat, 3), "html", digits = 4),
              bootstrap_options=c("striped", "hover", "condensed", "responsive"),
              full_width=FALSE) %>% kable_classic("hover", full_width = T) 
```
