---
title: "Analysis with split data"
author: "Ofer Dotan and Mariana Saldarriaga"
date: "2021"
output: html_document
---

#Install packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Load libraries
```{r packages, fig.show='hide', message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(grf)
library(haven)
library(caret)
library(tidyverse) 
library(expss)
library(base)
library(readr)
library(stats)
```

#Load data
```{r}
load("df_uganda.RData")
```

#Split the dataset into train and test set. Why? to not have overfiting problems. 
```{r}
set.seed(1236)
options(na.action = 'na.pass')
N <- nrow(df_uganda)
target <- round(N * 0.75)
rv <- runif(N)
df_uganda_train <- df_uganda[rv < 0.75, ]
df_uganda_test <- data.frame(df_uganda[rv >= 0.75, ])
```

#Generate X, Y, W (treatment), C(clusters) matrixes
```{r, include=FALSE}

## INCOME GENERATING ACTIVITY
Y.outcome <- as.matrix(df_uganda$post_income_gen_act)


#Baseline variables
X.vars <- stats::model.matrix(~ children + 
                            branchno + 
                            age_quant + 
                            rural + 
                            income +
                            below16+
                            rich+
                            club_frequently + 
                            club_sometimes + 
                            partner +
                            empl+
                            enroll_school + 
                            study_hours_quant +
                            worry_job +
                            work_married+
                            ablework_married+
                            want_respect+
                            who_decidehusband+
                            index_aspiration_quant+
                            index_iga_quant+
                            index_empowerment_quant+
                            satisf_income,
                            data= df_uganda)

#cluster
C.vill.id <- c(df_uganda$vill_id)

# Treatment assignment
W.treatment <- as.numeric(levels(df_uganda$treatment)[as.integer(df_uganda$treatment)])


```

# Estimate treatment effects with Causal Forests
train the Y.forest and Z.forest with default settings (later inputs to the causal forest)
```{r  message=FALSE, warning=FALSE, include=FALSE}

Y.forest = regression_forest(X.vars, 
                             Y.outcome, 
                             clusters = C.vill.id, 
                             num.trees = 6000,
                             equalize.cluster.weights = TRUE,
                             tune.parameters = "all")

Y.hat = predict(Y.forest)$predictions

W.forest = regression_forest(X.vars, 
                             W.treatment, 
                             clusters = C.vill.id,
                             num.trees = 6000,
                             equalize.cluster.weights = TRUE,
                             tune.parameters = "all")

W.hat = predict(W.forest)$predictions
```

#Train raw causal forest (no important variable selection)
```{r  message=FALSE, warning=FALSE, include=FALSE}
cf.raw = causal_forest (X.vars, 
                        Y.outcome, 
                        W.treatment,
                        num.trees = 6000,
                        Y.hat = Y.hat, 
                        W.hat = W.hat,
                        clusters = C.vill.id,
                        equalize.cluster.weights = TRUE,
                        tune.parameters = "all",
                        orthog.boosting = TRUE)


# Test set --> This is not necessarily the test set; just the x matrix in the Causal Forest model
options(na.action='na.pass')
X.test <- model.matrix( ~ children + 
                            branchno + 
                            age + 
                            rural + 
                            income +
                            rich+
                            club_frequently + 
                            club_sometimes + 
                            partner +
                            z_empl+
                            enroll_school + 
                            study_hours +
                            index_aspiration+
                            index_income_gen_act+
                            index_empowerment +
                            satisf_income,
                             data = df_uganda_test)

```


# Validating the overlap assumption.
Estimate the data's propensity scores to ensure scores are bounded away from 0 and 1
In order for conditional average treatment effects to be properly identified
```{r}
propensity.forest = regression_forest(X.vars, W.treatment)
W.hat.p = predict(propensity.forest)$predictions
hist(W.hat.p, xlab = "propensity score", main= paste("overlap assumption test: control over body index"))
```

# Estimate the average treatment effect for the raw cf
```{r}
# Average treatment effect for the raw model
ATE = average_treatment_effect(cf.raw)
ATE
paste("95% CI for the ATE:", round(ATE[1], 3),
      "+/-", round(qnorm(0.975) * ATE[2], 3))

```

# Find important variables
```{r}
varimp = variable_importance(cf.raw)
selected.idx = which(varimp > mean(varimp>0.2)) 
```

# Train a second causal forest using only important variables index
```{r}
#train a causal forest only with selected important variables.
cf = causal_forest(X.vars[,selected.idx], 
                      Y.outcome, W.treatment, 
                      Y.hat = Y.hat, 
                      W.hat = W.hat, 
                      num.trees = 6000,
                      clusters = C.vill.id,  
                      orthog.boosting = TRUE,
                      equalize.cluster.weights = TRUE,
                      tune.parameters = "all")
```


# Predict the test set and plot histogram of frequency of predictions
```{r}
#predict test set
tau.hat <- predict(cf, X.test[,selected.idx], 
                   num.trees = 6000, 
                   data=df_uganda_test, 
                   tune.parameters = "all",
                   estimate.variance = TRUE)

#plot frequency of predictions
hist((tau.hat)$predictions, main="Histogram of causal forest predictions (test sample)",
     xlab = "predictions")  
```


Plot predictions on test set
```{r}
# Estimate treatment effects for the test sample: NOT WORKING PLOT
plot(X.test[,selected.idx], tau.hat$predictions, xlab = "x", ylab = "tau", type = "l")
lines(X.test, pmax(0, X.vars), col = 2, lty = 2)
```
# Estimate ATE for cf model with important variables
```{r}
ATE.cf = average_treatment_effect(cf)
ATE.cf

paste("95% CI for the ATE:", round(ATE.cf[1], 3),
      "+/-", round(qnorm(0.975) * ATE.cf[2], 3))
```


#Check overlap assumption for selected index
#assumption holds when choosing only selected index
```{r}
propensity.forest.selected.inx = regression_forest(X.vars[,selected.idx], W.treatment)
W.hat.p = predict(propensity.forest.selected.inx)$predictions
hist(W.hat.p, main = "overlap assumption check: index empowerment", xlab = "propensity score")
```

# Estimate standard errors for test data and plot them
```{r}
# Standard error predictions
standard.error = sqrt(tau.hat$variance.estimates)
hist(standard.error) # This is the sigma part
```

# Alternative method to estimate standard errors test data
```{r}
model_sigma <- sqrt(predict(cf, X.test[, selected.idx], estimate.variance = TRUE)$variance.estimates)
```

```{r}
# Add confidence intervals for heterogeneous treatment effects (OJO: growing more trees is now recommended)
plot(X.test[, 1], tau.hat, ylim = range(tau.hat + 1.96 * sigma.hat, tau.hat - 1.96 * sigma.hat, 0, 2), xlab = "x", ylab = "tau", type = "l")
lines(X.test[, 1], tau.hat + 1.96 * sigma.hat, col = 1, lty = 2)
lines(X.test[, 1], tau.hat - 1.96 * sigma.hat, col = 1, lty = 2)
lines(X.test[, 1], pmax(0, X.test[, 1]), col = 2, lty = 1)
```

# Estimate CATE and plot estimations
```{r}
X.test.cate <- df_uganda_test %>% select(children,
                            branchno,
                            age, 
                            rural,
                            income,
                            rich,
                            club_frequently, 
                            club_sometimes, 
                            partner,
                            z_empl,
                            enroll_school,
                            study_hours,
                            index_aspiration,
                            index_income_gen_act,
                            index_empowerment,
                            satisf_income)

  
new_data_cates <- as.data.frame(X.test.cate[,-1])
new_data_cates$pred_est <- c(tau.hat$predictions)
new_data_cates$pred_var <- c(model_sigma)
new_data_cates$pred_est_lb <- new_data_cates$pred_est - 1.96 * new_data_cates$pred_var
new_data_cates$pred_est_ub <- new_data_cates$pred_est + 1.96 * new_data_cates$pred_var

# Get results for every level of every variable by aggregating up
cates <- lapply(names(new_data_cates[, 1:4]), function(x) {
  tmp <- new_data_cates %>% 
    group_by_(x) %>% 
    transmute(
      variable = x,
      ate = round(mean(pred_est) * 100, 2),
      ate_lb = round(mean(pred_est_lb) * 100, 2),
      ate_ub = round(mean(pred_est_ub) * 100, 2)
      )%>% 
    unique() %>% 
    as.data.frame()
  tmp <- tmp[, c(2, 1, 3, 4, 5)]
  names(tmp)[2] <- "level"
  tmp
})


cates <- do.call(rbind, cates) %>% 
  mutate_if(is.character, as.factor)

#cates.new.df <- as.data.frame(cates.new)

# visualize these
plot.cates <- ggplot(cates, aes(x = level, y = ate, colour = variable)) + 
  geom_point() +
  geom_errorbar(aes(ymin = ate_lb, ymax = ate_ub), width = .2) +
  geom_hline(yintercept = 0, linetype = 3) +
  theme_light() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.text.y = element_text(colour = "black"),
    strip.background = element_rect(colour = NA, fill = NA),
    legend.position = "none"
    ) + 
  facet_grid(variable ~ ., scales = "free_y") +
  coord_flip()

plot.cates  
```

# Estimate the overlap assumption to check ifcovariates are balanced across the treated and control group by plotting the inverse-propensity weighted histograms
```{r}
  
plot.df <- data.frame(X.vars, W = as.factor(W.treatment), IPW = ifelse(W.treatment == 1, 1 / W.hat, 1 / (1 - tau.hat)))

# on train or test??
r <- 23
plot.df.long <- reshape(plot.df, varying = list(1:r), v.names = "X", direction = "long",
                   times = factor(paste0("X.", 1:r), levels = paste0("X.", 1:r)))

ggplot(plot.df.long, aes(x = X.vars, weight = IPW, fill = W.treatment)) + 
  geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
  facet_wrap( ~ time, ncol = 2)

ggplot(plot.df, aes(x = X.vars, weight = IPW, fill = W.treatment)) 

dim(X.vars)
```

#Best linear predictor analysis to understand if the model detects heterogeneity 
```{r}
test_calibration(cf)
```

#Compare villages with high or low ATE estimations
```{r}
high_effect = tau.hat$predictions > median(tau.hat$predictions)
ate.high = average_treatment_effect(cf, subset = high_effect)
ate.low = average_treatment_effect(cf, subset = !high_effect)
paste("95% CI for difference in ATE:",
      round(ate.high[1] - ate.low[1], 3), "+/-",
      round(qnorm(0.975) * sqrt(ate.high[2]^2 + ate.low[2]^2), 3))
```

#Estimate model bias
```{r}
p <- mean(W.treatment)
Y.hat.0 <- cf$Y.hat - W.hat * tau.hat
Y.hat.1 <- cf$Y.hat + (1 - W.hat) * tau.hat

bias <- (W.hat - p) * (p * (Y.hat.0 - mean(Y.hat.0))  + (1 - p) * (Y.hat.1 - mean(Y.hat.1)))

hist(bias / sd(Y.outcome))
```

#Plot a tree
```{r}
library(DiagrammeR)
# Graph Causal Forests
plot(tree <- get_tree(cf, 3))
tree$nodes
```

# Run specific analysis to check heterogeneity in important variables
```{r}
# Formal test for specific variables
X.vars.df <- as.data.frame(X.vars)
dr.score = tau.hat$predictions + W.treatment / cf$W.hat *
  (Y.outcome - cf$Y.hat - (1 - cf$W.hat) * tau.hat$predictions) -
  (1 - W.treatment) / (1 - cf$W.hat) * (Y.outcome - cf$Y.hat + cf$W.hat * tau.hat$predictions) # propensity score weighting

# Village (cluster)
vill.id <- as.numeric(df_uganda$vill_id)
df_uganda$vill_id <- as.factor(df_uganda$vill_id)
options(na.action='na.pass')
vill.mat = model.matrix(~ vill_id + 0, data = df_uganda)
vill.size = colSums(vill.mat)
vill.score = t(vill.mat) %*% dr.score/vill.size

```

```{r}
#age
age <- X.vars.df$age
age <- as.numeric(age)

age.high = t(vill.mat) %*% (dr.score * (as.numeric(age)>= 16)) /
  t(vill.mat) %*% (as.numeric(age)>= 16)

age.low = t(vill.mat) %*% (dr.score * (X.vars$age < 16)) /
  t(vill.mat) %*% (X.vars$age < 16)

plot(age.high, age.low)
t.test(age.high-age.low)


#income
income.high = t(vill.mat) %*% (dr.score * (X.vars.df$income >= 36297.58)) /
  t(vill.mat) %*% (X.vars.df$income >= 36297.58)

income.low = t(vill.mat) %*% (dr.score * (X.vars$income < 36297.58)) /
  t(vill.mat) %*% (X.vars$income < 36297.58)

plot(income.high, income.low)
t.test(income.high - income.low)

#club distance
mean(df_uganda$dist_nearclub)
dist.far = t(vill.mat) %*% (dr.score * (X.vars$dist_nearclub >= 0.5484732)) /
  t(vill.mat) %*% (X.vars$dist_nearclub >= 0.5484732)

dist.near = t(vill.mat) %*% (dr.score * (X.vars$dist_nearclub < 0.5484732)) /
  t(vill.mat) %*% (X.vars$dist_nearclub < 0.5484732)

plot(dist.far, dist.near)
t.test(dist.far - dist.near)
```

is this wrong because of mixing up variables with diverging levels??
```{r}
# Rural (tested in original paper)
vill.rural = t(vill.mat) %*% X.vars$rural / vill.size
high.rural = vill.rural > median(vill.rural)
t.test(vill.score[high.rural], vill.score[!high.rural]) 

# Income (continuos) (tested in original paper)
vill.income = t(vill.mat) %*% X.vars$income / vill.size
high.income = vill.income > median(vill.income)
t.test(vill.score[high.income], vill.score[!high.income])

# Rich
vill.rich = t(vill.mat) %*% X.vars$rich / vill.size
high.income = vill.rich > median(vill.rich)
t.test(vill.score[high.income], vill.score[!high.income])

# Age - Below 16 (tested in original paper)
vill.age = t(vill.mat) %*% as.matrix(X.vars$age)
high.age = vill.age > median(vill.age)
t.test(vill.score[high.age], vill.score[!high.age])

# Club distance
vill.dist = t(vill.mat) %*% as.matrix(X.vars$dist_nearclub)
high.dist = vill.dist > median(vill.dist)
t.test(vill.score[high.dist], vill.score[!high.dist])
```

```{r}
# CATE by village 
ord <- order(order(Y.hat))
vill.id <- as.numeric(df_uganda$vill_id)
village.sort <- ord[vill.id]
```
